---
title: "Reproducible Research. Concepts and ideas"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Replication
The ultimate standard for strengthening scientific evidence is replication of findings and conducting studies with independent

- Investigators	
- Data
- Analytical methods
- Laboratories
- Instruments

Replication is particularly important in studies that can impact broad policy or regulatory decisions

## What’s Wrong with Replication?
Some studies cannot be replicated

- No time, opportunistic
- No money
- Unique

Reproducible Research: Make analytic data and code available so that others may reproduce findings

## How Can We Bridge the Gap?
Between **replication** and **nothing** we have **reproduciblity**

## Why Do We Need Reproducible Research?
New technologies increasing data collection throughput; data are more complex and extremely high dimensional

Existing databases can be merged into new “megadatabases”
Computing power is greatly increased, allowing more sophisticated analyses

For every field “X” there is a field “Computational X”

## Example: Reproducible Air Pollution and Health Research
Estimating small (but important) health effects in the presence of much stronger signals

Results inform substantial policy decisions, affect many stakeholders

> EPA regulations can cost billions of dollars

Complex statistical methods are needed and subjected to intense scrutiny

## Research Pipeline

```{r}

```

