---
title: "Exploring NOAA Storm Database"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## "The most harmful eventypes in health and economics"

## Synopsis
This analysis want to ask two questions, the first one is which types of event are the most harmful with respect to population health. The second one is see which types of events have the greatest economic consequences.

We explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

**More info about the data:**

- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

Know which events are the most harmful in heath and economics can support public policies.

## Data Processing
Load needed libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
```

Download and unzip data
```{r}
# Downloading data
file <- "repdata_data_StormData.csv.bz2"
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

# Check if the file exists
if(!file.exists(file)){
  download.file(fileURL,file, method="curl") 
}

# Check if you unzip the data previously
if(!file.exists(file)){
  unzip(file)
}

```

Reading data
```{r, cache=TRUE}
if(!exists("StormData")){
    StormData <- read.csv("~/datasciencecoursera/04_Reproducible_Research/course_project_2/repdata_data_StormData.csv", strip.white=TRUE)
}
```

Check structure and variable classes
```{r}
str(StormData)
```

Change some variables to correct class
```{r}
# BGN_DATE to date
StormData$BGN_DATE <- mdy_hms(StormData$BGN_DATE)

# Some variables to factor
StormData$COUNTYNAME <- as.factor(StormData$COUNTYNAME)
StormData$STATE <- as.factor(StormData$STATE)
StormData$EVTYPE <- as.factor(StormData$EVTYPE)
StormData$PROPDMGEXP <- as.factor(StormData$PROPDMGEXP)
StormData$CROPDMGEXP <- as.factor(StormData$CROPDMGEXP)
```

We need also standardize `EVTYPE` and `damage exp` in capital letters to deduplicate data.
```{r}
StormData <- StormData %>%
  mutate(EVTYPE=toupper(EVTYPE)) %>%
  mutate(PROPDMGEXP = toupper(PROPDMGEXP))  %>%
  mutate(CROPDMGEXP = toupper(CROPDMGEXP))
```

In order to calculate question 2, we need to change the damage exponential for numerical values. First we see how many elements we have is these variables.
```{r}
unique(StormData$PROPDMGEXP)
unique(StormData$CROPDMGEXP)
```

Change H, M, K, B for corresponding valor
```{r}
StormData$PROPDMGEXP <- gsub("B","1000000000",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("K","1000",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("M","1000000",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("H","100",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("+","0",StormData$PROPDMGEXP, fixed = TRUE)
StormData$PROPDMGEXP <- gsub("?","0",StormData$PROPDMGEXP, fixed = TRUE)
StormData$PROPDMGEXP <- gsub("-","0",StormData$PROPDMGEXP, fixed = TRUE)
StormData$PROPDMGEXP <- gsub("^$","0",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("5","100000",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("6","1000000",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("4","10000",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("2","100",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("3","10000000",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("1","10",StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- gsub("8","100000000",StormData$PROPDMGEXP)

StormData$CROPDMGEXP <- gsub("B","1000000000",StormData$CROPDMGEXP)
StormData$CROPDMGEXP <- gsub("K","1000",StormData$CROPDMGEXP)
StormData$CROPDMGEXP <- gsub("M","1000000",StormData$CROPDMGEXP)
StormData$CROPDMGEXP <- gsub("H","100",StormData$CROPDMGEXP)
StormData$CROPDMGEXP <- gsub("+","0",StormData$CROPDMGEXP, fixed = TRUE)
StormData$CROPDMGEXP <- gsub("?","0",StormData$CROPDMGEXP, fixed = TRUE)
StormData$CROPDMGEXP <- gsub("-","0",StormData$CROPDMGEXP, fixed = TRUE)
StormData$CROPDMGEXP <- gsub("^$","0",StormData$CROPDMGEXP)
StormData$CROPDMGEXP <- gsub("2","100",StormData$CROPDMGEXP)
```

Check againg
```{r}
unique(StormData$PROPDMGEXP)
unique(StormData$CROPDMGEXP)
```

## Results

### Question 1
**Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?**

We need to calculate the total **injuries** and **fatalities** by each type of event to determine the most harmful.

We group the dataset by event type and sum the total of injuries and fatalities.

```{r}
HarmfulData <- StormData %>%
  group_by(EVTYPE) %>%
  summarise(FATALITIES = sum(FATALITIES),
            INJURIES = sum(INJURIES)) %>%
  mutate(total_harmful = FATALITIES + INJURIES) %>% 
  arrange(desc(total_harmful)) %>% 
  slice(1:5)
HarmfulData
```

Plotting to visualize the most eventype harmful
```{r}
# Making copy to reorganize data
total_harmful <- HarmfulData
total_harmful$total_harmful <- NULL

# Reorganize data
final_harm <- gather(total_harmful, type, count, -EVTYPE)
# Dropping empty levels
final_harm$EVTYPE <- as.factor(final_harm$EVTYPE)
final_harm$EVTYPE <- droplevels(final_harm$EVTYPE)

# Plotting barplot
ggplot(final_harm, aes(x = reorder(EVTYPE, -count), y = count, fill = type)) + 
  geom_bar(stat = "identity") +
    labs(title = "Top Harmful EvenTypes",
       caption = "The staked bar chart shows the most harmful eventypes to population healh by fatalities and injuries.") +
  xlab("Even Types") + ylab("Harmful")
```

> From the above data the most harmful event to population health across United States is `TORNADO` with `96979` deaths or injuries.

### Question 2
**Across the United States, which types of events have the greatest economic consequences?**

First we select only the values on EXP more than 0
```{r}
total_damage <- StormData %>%
  select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
  filter(PROPDMGEXP > 0) %>%
  filter(CROPDMGEXP > 0)
```

Calculate total damage per PROP and CROP and calculate total damage in millions.
```{r}
total_damage$EVTYPE <- as.factor(total_damage$EVTYPE)
total_damage <- total_damage %>%
  mutate(prop_damage = PROPDMG * as.numeric(PROPDMGEXP)) %>%
  mutate(crop_damage = CROPDMG * as.numeric(CROPDMGEXP)) %>%
  mutate(prop_crop = (prop_damage + crop_damage)/1000000)
```

Group by event type and calculate total damage by event type, sorting and getting top 5.
```{r}
total_damage <- total_damage %>%  
  group_by(EVTYPE) %>%
  summarise(total = sum(prop_crop)) %>%
  arrange(desc(total)) %>% 
  slice(1:5)
```

Plotting total damage by event type
```{r}
# Dropping empty levels
total_damage$EVTYPE <- droplevels(total_damage$EVTYPE)
# Plotting barplot
ggplot(total_damage, aes(x = reorder(EVTYPE, -total), y = total)) + 
  geom_bar(stat = "identity") +
  labs(title = "Top Damage by EvenTypes",
       caption = "The barplot shows the event types that cause the most economic damage across de United States.") +
  xlab("Even Types") + ylab("Total Damage in millions")
total_damage
```

> From the above data the events that have the greatest economic consequences across United States are `FLOOD`, `HURRICANE/TYPHOON` and `TORNADO`.